---
title: "How to make world maps in R"
author: "By Rike Bolam"
date: "Last updated 15/07/2020"
output: 
  html_document:
    toc: yes
    toc_float: true
    toc_depth: 5
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

A short tutorial on plotting world maps in R.  

## Install and load packages

You will need rworldmap and ggplot2, which is part of the tidyverse package. We will plot some gapminder data, so will need that package too:

```{r }
library(rworldmap)
library(tidyverse)
library(gapminder)
```

## Load map data

This function will give you the data for plotting each country:

```{r }
map.all <- map_data(map = "world")
head(map.all) ## region denotes the country
```

## Get your own data in order

Your data need to match the country names in `map.all`. This is probably what will take you longest to sort out, but we will use some good enough data today:


```{r message = FALSE, warning = FALSE}

head(gapminder) ## This is the gapminder dataset
gapminder <- rename(gapminder, region = country) ## we need to rename column country to region, so we can merge gapminder and map.all
```

## Merge datasets

Then merge the two dataframes `map.all` and `all`:
```{r }
map.all <- left_join(map.all, gapminder, by = "region") 
head(map.all)

```

## Now you can make the plot:

```{r}
ggplot() + geom_map(data = map.all, map = map.all, 
                    aes(map_id = region,             ## map_id is the variable name of the countries,
                        x = long, y = lat, 
                        fill = pop)) +               ## fill will colour in each country
                                                     ## Use colour = varname to change colour of the outline
  scale_fill_gradientn(colours=terrain.colors(5)) +  ## this will change the colours in the gradient
  labs(fill = "Population size") +                   ## changes the name of the legend
  theme_classic() +                                  ## change the overall look of the plot
  theme(panel.background = element_rect(colour = "black", fill = "NA"), 
        axis.ticks.length = unit(.05, "cm"), 
        text = element_text(size = 16.5))            ## change some more of the defaults
```


